%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% EPFL report package, main thesis file
% Goal: provide formatting for theses and project reports
% Author: Mathias Payer <mathias.payer@epfl.ch>
%
% To avoid any implication, this template is released into the
% public domain / CC0, whatever is most convenient for the author
% using this template.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[a4paper,11pt,oneside]{report}
% Options: MScThesis, BScThesis, MScProject, BScProject
\usepackage[MScThesis,lablogo]{EPFLreport}
\usepackage{xspace}

\title{Adaptive Calibration Algorithms\\for Tunable Analog Peripherals}
\author{Khalil M'hirsi}
\supervisor{The Doctoral Student}
\adviser{Prof. Mathieu Salzmann}
%\coadviser{Second Adviser}
\expert{The External Reviewer}

\newcommand{\sysname}{FooSystem\xspace}

\begin{document}
\maketitle
\makededication
\makeacks

\begin{abstract}
Continuous depth-sensing in modern analog peripherals offers unprecedented hardware fidelity, yet current implementations rely on static, user-defined thresholds that ignore individual biomechanical variance. Factors such as resting tremors, micro-hesitations, and physiological fatigue create "mechanical friction" where hardware sensitivity outpaces user stability, leading to unintended actuations and sub-optimal latency. To address this, we propose a personalized, data-driven system that dynamically tunes Actuation Level (AL) and Rapid Trigger (RT) parameters by analyzing raw telemetry, including trajectory, velocity, and resting pressure. By leveraging statistical profiling of professional input signatures and a low-overhead TinyML pipeline, our system enables real-time intent detection and threshold adaptation. Evaluation demonstrates that this model minimizes mechanical latency and mitigates the impact of fatigue without introducing computational system lag.
\end{abstract}

\maketoc

%%%%%%%%%%%%%%%%%%%%%%
\chapter{Introduction}
%%%%%%%%%%%%%%%%%%%%%%

Human-Computer Interaction (HCI) in high-performance computing and competitive esports has consistently pushed the boundaries of peripheral design. In environments where users execute upwards of 500 Actions Per Minute (APM), hardware latency and actuation precision are critical bottlenecks to both performance and musculoskeletal health~\cite{esports_tendinopathy, wrist_fatigue_genre_2025}. 

\section{The Evolution of Continuous Sensing Hardware}
The landscape of human-computer interaction (HCI) is undergoing a paradigm shift driven by advancements in peripheral hardware, transitioning from traditional binary mechanical switches to continuous analog sensing. Historically, optimizing selection techniques and activation thresholds has been a foundational goal in HCI \cite{touchpad_comparison, impact2018}. However, recent breakthroughs in magnetic and inductive switches have revolutionized input granularity. The commercialization of Hall-effect keyboards featuring "Rapid Trigger" technology allows for dynamic keystroke actuation and reset points, granting users unprecedented control in time-critical environments ranging from competitive gaming \cite{wooting_rapid_trigger, pcgamer_hall_effect} to human-piloted drone racing \cite{drone_rapid_trigger}. Similarly, the introduction of the Haptic Inductive Trigger System (HITS), as exemplified by Logitech's Superstrike technology, redefines mouse inputs by merging precise inductive sensing with customizable haptic feedback \cite{logitech_superstrike_2026, logitech_hits_patent, fox_superstrike_2026, techradar_superstrike_2026}. Alongside firmware advancements like tap-hold macro integrations \cite{attackshark_taphold} and fine-tuned magnetic sensors \cite{magnetic_tuning}, these devices are bridging the gap between digital input and continuous physical expression, much like the dynamic finger-key interactions observed in piano performance \cite{piano_3d_force} and early dynamic keyboard concepts \cite{dynamic_keyboard}.

\section{The Limitations of Static Actuation Thresholds}
Despite these hardware leaps, the software interpreting this data remains limited. Current analog peripherals rely on static firmware thresholds that are agnostic to human biomechanical variance. Players exhibit unique control signatures, including micro-hesitations and varying velocity curves, which often clash with rigid hardware logic~\cite{impact2018}. When static thresholds are set too aggressively, they fail to filter out physiological noise; when set too conservatively, they introduce unnecessary mechanical pre-travel. This friction results in sub-optimal latency or a high frequency of unintended actuations during high-stress scenarios~\cite{touchpad_comparison, dynamic_keyboard}.

\section{Neuromotor Fatigue and Performance Decay}
Furthermore, these advancements in gaming hardware intersect with the physical well-being of the user. The exponential rise in actions-per-minute (APM) in esports has led to a parallel increase in repetitive strain injuries and tendinopathies \cite{esports_tendinopathy, difrancisco2019managing, 1hp_esports_ergonomics}. Biomechanical research highlights distinct fatigue development patterns between keyboard and mouse usage \cite{fatigue_flexor}, with repetitive clicking directly degrading aiming performance and inducing forearm muscle fatigue \cite{clicking_fatigue_ergonomics_2024}. In fact, high-APM gaming genres impose specific kinematic burdens on wrist extensors \cite{wrist_fatigue_genre_2025}, while the brain itself undergoes neuromotor compensation-sharpening finger-specific representations-during fatiguing motor tasks \cite{heimhofer2024finger}. By allowing for lighter, customizable actuation forces, next-generation analog switches present a vital mechanical intervention to mitigate these biomechanical degradations.

Beyond mechanical actuation, the continuous data stream generated by Hall-effect and inductive sensors unlocks profound capabilities for passive user profiling. The ability to measure precise force and key travel enables advanced keystroke dynamics for continuous, biometric authentication \cite{keystroke_dynamics_arxiv, continuous_keystroke_mdpi, continuous_auth_2021}. Modern machine learning architectures, such as Transformer models applied to keystroke biometrics (e.g., TypeFormer), demonstrate high efficacy in user verification \cite{typeformer2022}. Conversely, this rich analog data can be leveraged for health and cognitive monitoring. Variations in typing pressure and force have been successfully correlated with user stress levels \cite{stress_pressure, piezo_ann}, and dynamic input adaptation holds immense potential for accessibility and rehabilitation, such as evaluating motor signs in Parkinson’s disease \cite{rehabeelitation}.

\section{Adaptive Intent Detection via Edge AI}
To fully realize the potential of continuous input sensing, the ecosystem must move from reactive hardware to predictive intent detection. Surface electromyography (sEMG) represents the ultimate modality for this, capturing neuromuscular signals before physical actuation occurs. Recent advancements in sEMG processing utilize masked modeling \cite{reactemg}, improved LSTMs \cite{emg_lstm}, and LDA-based approaches \cite{lda_emg} for low-latency, real-time intent detection, bolstered by large-scale multi-user datasets enabling zero-shot gesture recognition \cite{bigdata_emg}. Synthesizing high-polling-rate analog switch data with complex EMG signals requires shifting computational workloads directly to the peripheral device. Consequently, the integration of Edge AI and TinyML architectures-specifically utilizing fault-tolerant RISC-V SoCs-is crucial for executing efficient, localized inference without introducing latency \cite{tinyml_mcu_inference, tinyml_edge_2025, singh2023edge}.

\section{Thesis and Contributions}
The core thesis of this work is that dynamic, real-time intent prediction can eliminate the friction caused by static thresholds. By analyzing continuous telemetry, we can extract a user's "control signature" and adjust RT and AL buffers on the fly to compensate for fatigue. The contributions of this project are:
\begin{itemize}
    \item A robust, low-latency telemetry processing architecture for continuous inductive and Hall-effect sensors.
    \item A suite of adaptive calibration algorithms that tune AL and RT parameters based on physiological state.
    \item A theoretical framework for fatigue detection using snap-back velocity and resting pressure changes.
    \item An evaluation of on-device TinyML inference on RISC-V architectures to demonstrate zero-lag threshold adaptation.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%
\chapter{Background}
%%%%%%%%%%%%%%%%%%%%%%

To understand the design and implementation of an adaptive actuation system, it is necessary to establish the technical foundations of continuous sensing hardware, the mathematical logic of modern trigger parameters, and the biomechanical constraints of the human hand.

\section{Continuous Depth-Sensing Mechanisms}

Traditional mechanical switches operate on a binary principle: a circuit is either open or closed based on a physical contact point. In contrast, continuous depth-sensing switches measure the absolute position of the stem throughout its entire travel distance.

\subsection{Hall-Effect and Magnetic Sensing}
Hall-effect sensors utilize a permanent magnet attached to the switch stem and a sensor on the Printed Circuit Board (PCB). As the magnet approaches the sensor, the Hall voltage varies proportionally to the magnetic field strength~\cite{wooting_rapid_trigger, pcgamer_hall_effect}. This allows the firmware to map voltage levels to precise displacement values in millimeters. This technology eliminated "debounce" delay-the physical vibrations of metal leaves-allowing for near-instantaneous signal processing.

\subsection{Haptic Inductive Trigger System (HITS)}
Refined in the 2026 Logitech SUPERSTRIKE series, the Haptic Inductive Trigger System (HITS) replaces magnets with an inductive coil architecture~\cite{logitech_hits_patent, techradar_superstrike_2026}. By measuring changes in electromagnetic induction caused by the proximity of a metallic puck within the trigger, HITS provides higher resolution and lower susceptibility to external magnetic interference compared to Hall-effect sensors. Furthermore, HITS integrates localized haptic motors to simulate tactile "bumps" or "clicks" at software-defined depths, decoupling the physical feel of the switch from its electrical actuation point~\cite{fox_superstrike_2026}.

\section{Control Parameters: AL and RT}

The transition to continuous sensing introduced two primary variables that define the "feel" and speed of an input device.

\subsection{Actuation Level (AL)}
The Actuation Level (AL) is the specific depth ($d$) at which a "key down" event is registered. In traditional switches, this is fixed (e.g., 2.0mm). In analog systems, AL is a variable $d_{act} \in [0.1, 4.0]$mm. Lowering $d_{act}$ reduces mechanical pre-travel, allowing for faster response times, but increases the risk of accidental actuations caused by resting finger weight.

\subsection{Rapid Trigger (RT)}
Rapid Trigger (RT) eliminates the fixed reset point found in mechanical switches. In a standard switch, the key must travel back up past a specific "reset point" before it can be pressed again. RT logic instead monitors the direction of travel. As soon as the stem moves upward by a specific threshold ($\Delta_{up}$), the input resets. Conversely, as soon as it moves back down by $\Delta_{down}$, it reactuates~\cite{wooting_rapid_trigger, drone_rapid_trigger}. This allows for hyper-fast "spamming" of inputs, which is critical in high-APM genres.

\section{Biomechanical Variance and Neuromotor Fatigue}

The efficiency of AL and RT is limited by the physiological state of the user. Human motor control is subject to "signal-dependent noise," where the variance of movement increases with the force or speed of the action.

\subsection{Resting Pressure and Tremors}
Even at rest, users exert varying degrees of force on a peripheral. Factors such as cognitive stress or caffeine intake can cause micro-tremors-high-frequency, low-amplitude oscillations in displacement telemetry~\cite{stress_pressure}. If the AL is set to 0.1mm and the user's tremor amplitude is 0.15mm, the system will register "ghost" inputs.

\subsection{Neuromotor Fatigue Signatures}
Extended sessions lead to fatigue in the \textit{flexor digitorum superficialis} and \textit{extensor digitorum} muscles~\cite{fatigue_flexor, clicking_fatigue_ergonomics_2024}. Fatigue manifests in two measurable telemetry changes:
\begin{enumerate}
    \item \textbf{Velocity Decay:} The "snap-back" velocity (the speed at which a finger releases a key) decreases as the extensor muscles tire~\cite{tholl_wrist_fatigue_2025}.
    \item \textbf{Resting Weight Shift:} As the flexor muscles lose tonicity, the resting weight of the finger often increases, leading to deeper "resting" displacement in the analog sensor~\cite{heimhofer2024finger, 1hp_esports_ergonomics}.
\end{enumerate}

\section{On-Device Inference and TinyML}

Processing continuous telemetry at high polling rates (e.g., 8,000Hz) requires low-latency computation that cannot be achieved through standard OS-level drivers due to USB polling intervals and kernel scheduling~\cite{monaco_sok_2018}.

\subsection{RISC-V and Vector Extensions}
Modern high-performance peripherals utilize 32-bit MCUs based on the RISC-V architecture. Recent iterations include vector extensions that allow for parallel processing of telemetry arrays~\cite{tinyml_mcu_inference}. This hardware is capable of running "TinyML", quantized machine learning models (such as Linear Discriminant Analysis or lightweight LSTMs) that have been optimized for minimal memory and power footprints~\cite{tinyml_edge_2025, singh2023edge}.

\subsection{Quantization and Latency}
To maintain a sub-millisecond pipeline, ML models must undergo quantization, converting 32-bit floating-point weights into 8-bit integers (INT8). This reduction allows the MCU to perform "pre-report" inference-predicting if a movement is an intended click or a fatigue-induced tremor before the HID (Human Interface Device) report is even sent to the computer.


%%%%%%%%%%%%%%%%
\chapter{Design}
%%%%%%%%%%%%%%%%

Introduce and discuss the design decisions that you made during this project.
Highlight why individual decisions are important and/or necessary. Discuss
how the design fits together.

This section is usually 5-10 pages.


%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Implementation}
%%%%%%%%%%%%%%%%%%%%%%%%

The implementation covers some of the implementation details of your project.
This is not intended to be a low level description of every line of code that
you wrote but covers the implementation aspects of the projects.

This section is usually 3-5 pages.


%%%%%%%%%%%%%%%%%%%%
\chapter{Evaluation}
%%%%%%%%%%%%%%%%%%%%

In the evaluation you convince the reader that your design works as intended.
Describe the evaluation setup, the designed experiments, and how the
experiments showcase the individual points you want to prove.

This section is usually 5-10 pages.


%%%%%%%%%%%%%%%%%%%%%%
\chapter{Related Work}
%%%%%%%%%%%%%%%%%%%%%%

The related work section covers closely related work. Here you can highlight
the related work, how it solved the problem, and why it solved a different
problem. Do not play down the importance of related work, all of these
systems have been published and evaluated! Say what is different and how
you overcome some of the weaknesses of related work by discussing the 
trade-offs. Stay positive!

This section is usually 3-5 pages.


%%%%%%%%%%%%%%%%%%%%
\chapter{Conclusion}
%%%%%%%%%%%%%%%%%%%%

In the conclusion you repeat the main result and finalize the discussion of
your project. Mention the core results and why as well as how your system
advances the status quo.

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\printbibliography

% Appendices are optional
% \appendix
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \chapter{How to make a transmogrifier}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% In case you ever need an (optional) appendix.
%
% You need the following items:
% \begin{itemize}
% \item A box
% \item Crayons
% \item A self-aware 5-year old
% \end{itemize}

\end{document}